## Amyloid Positivity Exploration

# Motivation/Goal
This project was created over the span of a week long ML hackathon held virtually at UCI. Hackathon End Date: 04/25/2020

Our main research question we focused our efforts on addressing was predicting which people tested Amyloid Positive based on other features in the dataset.

For context, Amyloid Beta, the cerebrospinal fluid measurement taken at the time of assessment (or within 6 months of assessment) are believed to be one of the first indicators of otherwise pre-symptomatic Alzheimer's disease. Amyloid Positivity diagnosis is then determined according to the evidence-based cutoff of 192 pg/mL. Amyloid Beta levels below 192 pg/mL is considered Amyloid Negative, while above 192 pg/mL is considered Amyloid Positive.

Some other topics of analysis we explored inlcluded predicting future test scores with past test scores, investigating the existance of different clusters or subgroups within the data, and thoroughly understanding the relationship between combinations of different features.

# Libraries
```{r}
library(aod)
library(caret)
library(ggplot2)
library(InformationValue)
library(psychotree)
library(smbinning)
library(tidyverse)
```

# Raw Data
Unprocessed data from 292 patients provided by the Alzheimer’s Disease Neuroimaging Initiative. There are 939 assessments represented in terms of 19 variables from the AVLTAmyloidClassification dataset.

Demographic:
* Age
* Sex
* Years of Education

Protocol Information:
* Time Since Baseline Test (Month 0)

Memory Test Outcomes:
* Free Recall Scores
* Recognition Scores

Biomarkers:
* APOE Genotype
* Beta Amyloid

Diagnosis of Memory Impairment:
* Cognitively Normal/Impaired

```{r}
amyloid <- read_csv('./Data/Amyloid.csv')
```

# Summary Statisitics
Our data at a glance. 

Continuous features are summarized as means, while the categorical features are summarized as proportions.
  
## A. Demographics:
```{r}
amyloid_demographic_baseline_summary <- amyloid %>%
  filter(month == 0) %>%
  summarise(
    mean_age  = mean(age), 
    mean_edu = mean(edu), 
    prop_male = sum(sex == 2)/n(),
    prop_female = sum(sex == 1)/n(),
  )
```
## B. Protocol Information:
```{r}
amyloid_test_performance_baseline_summary <- amyloid %>%
  filter(month == 0)  %>%
  group_by(abeta6mcut) %>% 
  summarise(
    mean_t1sum = mean(t1sum, na.rm = TRUE),
    mean_t2sum = mean(t2sum, na.rm = TRUE),
    mean_t3sum = mean(t3sum, na.rm = TRUE),
    mean_t4sum = mean(t4sum, na.rm = TRUE),
    mean_t5sum = mean(t5sum, na.rm = TRUE),
    mean_t6sum = mean(t6sum, na.rm = TRUE),
    mean_t7sum = mean(t7sum, na.rm = TRUE),
    drec_hits = mean(drec_hits, na.rm = TRUE),
    drec_fa = mean(drec_fa, na.rm = TRUE),
  )
```
## C. Memory Test Outcomes:
```{r}
amyloid_avlt_baseline_summary <- amyloid %>%
  filter(month == 0)  %>%
  group_by(abeta6mcut) %>% 
  summarise(
    mean_t1sum = mean(t1sum, na.rm = TRUE),
    mean_t2sum = mean(t2sum, na.rm = TRUE),
    mean_t3sum = mean(t3sum, na.rm = TRUE),
    mean_t4sum = mean(t4sum, na.rm = TRUE),
    mean_t5sum = mean(t5sum, na.rm = TRUE),
    mean_t6sum = mean(t6sum, na.rm = TRUE),
    mean_t7sum = mean(t7sum, na.rm = TRUE),
  )
```

## D. Biomarkers:
```{r}
amyloid_biomarker_baseline_summary <- amyloid %>%
  filter(month == 0) %>%
  summarise(
    mean_abeta6_measure = mean(abeta6m, na.rm),
    prop_positive = sum(abeta6mcut == 1)/n(),
    prop_negative = sum(abeta6mcut == 2)/n(),
    prop_e2e2 = sum(genotype == 1)/n(),
    prop_e2e3 = sum(genotype == 2)/n(),
    prop_e3e3 = sum(genotype == 3)/n(),
    prop_e2e4 = sum(genotype == 4)/n(),
    prop_e3e4 = sum(genotype == 5)/n(),
    prop_e4e4 = sum(genotype == 6)/n()
  )
```

## E. Diagnosis of Memory Impairment
```{r}
amyloid_diagnosis_baseline_summary <- amyloid %>%
  filter(month == 0) %>%
  summarise(
    prop_normal_cognition = sum(dx == 1)/n(),
    prop_subjectively_impaired = sum(dx == 2)/n(),
    prop_objectively_impaired = sum(dx == 3)/n(),
  )
```


# Preprocessing

For preprocessing we:
1. Factorized categorical features  
2. Replaced all instances of NA (represented as negative quantities) with 0
3. Added the recognition test proportion (drec_hits / drec_fa)
4. Added the difference in test scores for select AVLT intervals
    + Learning Effect over immediate recall (t1t5)
    + Delayed Recall Effect (t5t6, t5t7)
    + Relative Improvement or Decline From Previous Test *i* (t*i*-t(*i*-1))

```{r}
amyloid <- amyloid %>%
  mutate(
    month = as.factor(if_else(month < 0, NA_real_, month)),
    abeta6m = if_else(abeta6m < 0, NA_real_, abeta6m),
    abeta6mcut = factor(abeta6mcut),
    sex = factor(sex),
    genotype = factor(genotype),
    dx = factor(dx),
    edu = if_else(edu < 0, NA_real_, edu),
    wordlist = factor(wordlist),
    t1sum = if_else(t1sum < 0, NA_real_, t1sum),
    t2sum = if_else(t2sum < 0, NA_real_, t2sum),
    t3sum = if_else(t3sum < 0, NA_real_, t3sum),
    t4sum = if_else(t4sum < 0, NA_real_, t4sum),
    t5sum = if_else(t5sum < 0, NA_real_, t5sum),
    t6sum = if_else(t6sum < 0, NA_real_, t6sum),
    t7sum = if_else(t7sum < 0, NA_real_, t7sum),
    distsum = if_else(distsum < 0, NA_real_, distsum),
    drec_hits = if_else(drec_hits < 0, NA_real_, drec_hits),
    drec_fa = if_else(drec_fa < 0, NA_real_, drec_fa),
    t1t2 = t2sum - t1sum,
    t2t3 = t3sum - t2sum,
    t3t4 = t4sum - t3sum,
    t4t5 = t5sum - t4sum,
    t5t6 = t6sum - t5sum,
    t6t7 = t7sum - t6sum,
    t1t5 = t5sum - t1sum,
    t5t7 = t7sum - t5sum,
    recognition_prop = drec_hits/drec_fa
  )


  baseline <- amyloid %>% filter(month == 0)
```

# Exploratory Data Analysis (EDA)

## Exploring the association of Demographic/Biomarker/Diagnosis features with Performance Measures and Amyloid Beta Measures .
 #### Key Observations
 * Genotype was strongly associated with baseline AVLT scores and Beta Amyloid measures
 * There was a strong positive relationship between sex and baseline AVLT scores (females tend to score better), however, Amyloid Beta measures do not seem to vary by sex 
 * Education did not seem to be associated with either AVLT scores or Amyloid Beta levels
 * Neither test performance nor amyloid beta levels seemed to vary 

 #### Inference
 * Genotype is a discriminatory factor when predicting Beta Amyloid
 * Because sex is correlated with AVLT scores, but not Amyloid Beta levels, including sex into a linear model would introduce multicollinearity
        
### A. Genotype
```{r}
par(mfrow = c(1,2))
baseline %>% gather("test_number", "score", c(t1sum, t2sum, t3sum, t4sum, t5sum, t6sum, t7sum)) %>%
  ggplot(aes(x = test_number, fill = genotype)) +
  scale_x_discrete(labels = c("IR 1", "IR 2", "IR 3", "IR 4", "IR 5", "DR 1", "DR 2")) +
  geom_boxplot(aes(x= test_number, y = score)) +
  scale_fill_discrete(name = "Genotype", labels = c("e2/e2", "e2/e3", "e3/e3", "e2/e4", "e3/e4", "e4/e4")) +
  labs(x = "AVLT", y = "Score") +
  ggtitle("AVLT Scores by Genotype")

amyloid %>% 
  ggplot(aes(x = month, fill = genotype)) +
  geom_boxplot(aes(x= month, y = abeta6m)) +
  scale_fill_discrete(name = "Genotype", labels = c("e2/e2", "e2/e3", "e3/e3", "e2/e4", "e3/e4", "e4/e4")) +
  labs(x = "Month", y = "Amyloid Beta Levels") + 
  ggtitle("Amyloid Beta Measures vs Time, by Genotype")
par(mfrow = c(1,1))
```

### B. Sex
```{r}
par(mfrow = c(1,2))
baseline %>% gather("test_number", "score", c(t1sum, t2sum, t3sum, t4sum, t5sum, t6sum, t7sum)) %>%
  ggplot(aes(x = test_number, fill = sex)) +
  scale_x_discrete(labels = c("IR 1", "IR 2", "IR 3", "IR 4", "IR 5", "DR 1", "DR 2")) +
  geom_boxplot(aes(x= test_number, y = score)) +
  scale_fill_discrete(name = "Sex", labels = c("Female", "Male")) +
  labs(x = "AVLT", y = "Score") + 
  ggtitle("AVLT Scores by Sex")

amyloid %>%
  ggplot(aes(x = month, fill = sex)) +
  geom_boxplot(aes(x= month, y = abeta6m)) +
  scale_fill_discrete(name = "Sex", labels = c("Female", "Male")) +
  labs(x = "Month", y = "Amyloid Beta Levels") + 
  ggtitle("Amyloid Beta Measures vs Time, by Sex")
par(mfrow = c(1,1))
```

### C. Education
```{r}
par(mfrow = c(1,2))
baseline %>% gather("test_number", "score", c(t1sum, t2sum, t3sum, t4sum, t5sum, t6sum, t7sum)) %>% 
  mutate(edu_factor = as.factor(ntile(edu, 3))) %>%
  ggplot(aes(x = test_number, fill = edu_factor)) +
  scale_x_discrete(labels = c("IR 1", "IR 2", "IR 3", "IR 4", "IR 5", "DR 1", "DR 2")) +
  geom_boxplot(aes(x= test_number, y = score)) +
  scale_fill_discrete(name = "Education Bracket", labels = c("<= Bachelors", "<= Masters", "<= PhD")) +
  labs(x = "AVLT", y = "Score") + 
  ggtitle("AVLT Scores by Education Group")

amyloid %>%
  mutate(edu_factor = as.factor(ntile(edu, 3))) %>%
  ggplot(aes(x = month, fill = edu_factor)) +
  geom_boxplot(aes(x = month, y = abeta6m)) +
  scale_fill_discrete(name = "Diagnosis", labels = c("<= Bachelors", "<= Masters", "<= PhD")) +
  labs(x = "Month", y = "Amyloid Beta Levels") +
  ggtitle("Amyloid Beta Measures vs Time, by Education")
par(mfrow = c(1,1))
```

### D. Age
```{r}
par(mfrow = c(1, 2))
baseline %>% gather("test_number", "score", c(t1sum, t2sum, t3sum, t4sum, t5sum, t6sum, t7sum)) %>% 
  mutate(age_factor = as.factor(ntile(age, 3))) %>%
  ggplot(aes(x = test_number, fill = age_factor)) +
  scale_x_discrete(labels = c("IR 1", "IR 2", "IR 3", "IR 4", "IR 5", "DR 1", "DR 2")) +
  geom_boxplot(aes(x= test_number, y = score)) +
  scale_fill_discrete(name = "Age Range", labels = c("[54-72)", "[72-77)", "[77, 89]")) +
  labs(x = "AVLT", y = "Score") +
  ggtitle("AVLT Scores by Age Group")

amyloid %>%
  mutate(age_factor = as.factor(ntile(age, 3))) %>%
  ggplot(aes(x = month, fill = age_factor)) +
  geom_boxplot(aes(x= month, y = abeta6m)) +
  scale_fill_discrete(name = "Age", labels = c("[54-72)", "[72-77)", "[77, 89]")) +
  labs(x = "Month", y = "Amyloid Beta Levels") +
  ggtitle("Amyloid Beta Measures vs Time, by Age Group")
par(mfrow = c(1,1)
```

### E. Diagnosis
```{r}
par(mfrow = c(1, 2))
baseline %>% gather("test_number", "score", c(t1sum, t2sum, t3sum, t4sum, t5sum, t6sum, t7sum)) %>%
  ggplot(aes(x = test_number, fill = dx)) +
  scale_x_discrete(labels = c("IR 1", "IR 2", "IR 3", "IR 4", "IR 5", "DR 1", "DR 2")) +
  geom_boxplot(aes(x= test_number, y = score)) +
  scale_fill_discrete(name = "Diagnosis", labels = c("Cognitively Normal", "Subjectively Cognitively Impaired", "Objective Mild Cognitive Impairment")) +
  labs(x = "AVLT", y = "Score") +
  ggtitle("AVLT Scores by Diagnosis")
  
amyloid %>%
  ggplot(aes(x = month, fill = dx)) +
  geom_boxplot(aes(x= month, y = abeta6m)) +
  scale_fill_discrete(name = "Diagnosis", labels = c("Cognitively Normal", "Subjectively Cognitively Impaired", "Objective Mild Cognitive Impairment")) +
  labs(x = "Month", y = "Amyloid Beta Levels") +
  ggtitle("Amyloid Beta Measures vs Time, by Diagnosis")
par(mfrow = c(1,1)
```

## Exploring how various performance measures and their linear combinations are associated with Amyloid Positivity
#### Key Observations:
* Delta measure between scores on successive administrations of the immediate recall tasks did not vary much by Amyloid Positivity
* Overall change in performance across the immediate recall tasks (t5sum - t1sum) was noticeably different across Amyloid Positive and Negative Participants
* The ratio of hits to false alarms on the Recognition task seems to be greater for those who tested Amyloid Postive
   
#### Inference:
* The learning effect accross the immediate recall discriminates between Amyloid Postive and Negative participants far better than the individual test measures 
   
# Feature Engineering
-> Test Differences/Learning Scores
        t1t5 (based of of (1... the learning is what seemd to discriminate between +ve and -ve)
        reference observation 1

# Question 1: Cluster Participants based on Recognition Task Performance
    -> Identify the Contitive Process for the Recognition Task
    (show output of spec)
    -> Computing y matrix, which was the outcome for the model
    -> drec_hit, drec_fa used to compute all counts
    -> explain r, b
    -> show various trees, based on various categories
    -> train all feature model

    -> Discuss interesting trees 
    (maybe the abeta cutoff...and compare it to the actual cutoff)

    -> Age Cutoffs with Recognition Test:
      (-inf, 67.2], (67.2, 69.9], (69.9, +inf)
      the middle group performs worse than the other two

    -> Recall 
        MPT -> to FEATURE ENGINEER base off of thresholds computed
    
# Question 2: Predict Amyloid Positivity with Feature Selection
    -> Compute Corrolation matrix
    (highlight features that have high corrolation with abeta6m)
    -> Logistic Regression Model with select features
    (compare feature combinations to eachother)
    -> Train one with a feature based off of the MPT cutoff categories


# Question 3: Predict test score trajectory in a single visit? [Maybe]
    -> An idea we had: LME, LC
    (to predict test score across visits)

```{r}
print("hello")
```

TODO:
[X] Clean Preprosseing
[X] Clean EDA Viz
[X] Clean more EDA
[X] Merge eda with more eda
[X] pick intersting EDA
[X] clean up Logistic Regression Model with test_train_split
[X] clean up MPT.R
[ ] Add code and text explaintions to RMarkdown
[ ] Answer Question 3 with LC model
[ ] Train Logistic Regression with feature based off MPT cutoff
[ ] Record Presentation (Allocate 2 hours)

For Fun
[ ] Feature Selection 
[ ] Create ensemble with Logistic regression models trained on various visit times seperately.

PRESENTATION NOTES:
* mention only DISCRIMINTORY FACTORS observations
    * scroll over plots relevant to the observations we are mentioning
    but dont dwell

